### 2015-09-12 thesis-proposal/problem.tex
Updates per Bob's comments

\paragraph{Abstract Syntax} The abstract syntax of {patterns}, $p$, over strings, $s$, is specified as below:\[p ::= \textbf{empty} ~|~ \textbf{str}(s) ~|~ \textbf{seq}(p; p) ~|~ \textbf{or}(p; p) ~|~ \textbf{star}(p) ~|~ \textbf{group}(p)\]
One way to express this abstract syntax is by defining a recursive sum type \cite{pfpl}. Verse supports these as case types, which are comparable to datatypes in ML (we plan to show how the type structure  of case types can in fact be expressed in libraries later):

\begin{lstlisting}[numbers=none]
casetype Pattern {
    Empty | Str of string | Seq of Pattern * Pattern 
    | Or of Pattern * Pattern | Star of Pattern | Group of Pattern
}
\end{lstlisting}

@@

\paragraph{Formalization} To introduce the formal specification of TSMs, let us consider the rule in the EL's semantics for handling TSM application to a delimited form. To simplify matters, we will first consider only unparameterized TSMs (i.e. TSMs defined at a single type, e.g. the case type \lstinline{Pattern}, rather than a parameterized family of types): 
\begin{mathpar}
\inferrule[syn-aptsm]{
    \vdash \psi ~@~ \sigma_\text{ty}~\{\sigma_\text{parser}\}\\
    \mathbf{parsestream}(b) = \sigma_\text{ps}\\
    \sigma_\text{parser}(\sigma_\text{ps}) \Downarrow \sigma_\text{exp}\\
    \sigma_\text{exp} \uparrow e_\text{exp}\\\\
    \Gamma; \emptyset \vdash e_\text{exp} \Leftarrow \sigma_\text{ty} \leadsto \iota_\text{trans}
}{
    \Gamma_\text{out}; \Gamma \vdash \mathbf{aptsm}(\psi; b) \Rightarrow \sigma_\text{ty} \leadsto \iota_\text{trans}
}
\end{mathpar}
Looking at the conclusion of the rule, we see that TSM application takes the abstract form $\mathbf{aptsm}(\psi; b)$. The metavariable $\psi$ ranges over \emph{TSM expressions} and $b$ ranges over {bodies} of delimited forms (i.e. the sequences of characters between the delimiters in the examples above). TSM expressions consist only of TSM definitions here (cf. the definition of \lstinline{pattern} above), though when we formalize parameterized TSMs, we will distinguish TSM definitions from TSM expressions, which involve parameter application.%Note that for simplicity in the formal system, we have tacitly assumed that TSM names are drawn from a set that is disjoint from the set of variable names, so that function application and TSM application desugar to the distinct abstract forms. %In practice, we would use a common set of names and include a disambiguation step, but making the assumption of disjoint names allows us to avoid specifying this. %In practice, there would be an additional step that determined whether a function or a TSM was being applied based on the current context,

@@

As another example, we can use TSLs to express derived list syntax. For example, if we use a case type, we can define the TSL directly upon declaring the list type:
\begin{lstlisting}[numbers=none]
casetype list('a) { Nil | Cons of 'a * list('a) } with syntax {
  static fn (body :: ParseStream) => 
    (* ... comma-delimited spliced exps ... *)
}
\end{lstlisting}
Together with the TSL for patterns, this allows us to write a list of patterns like this:
\begin{lstlisting}[numbers=none]
let val x : list(P.t) = [/SURL\dEURL/SHTML, EHTML/SURL\d\dEURL/SHTML, EHTML/SURL\d\d\dEURL/]
\end{lstlisting}
From the client's perspective, it is essentially as if the language had built in derived syntax for lists and regular expression patterns directly.%However, we did not need to build in this syntax primitively.%The only constraint is that this syntax must be used in an analytic position, which we argue is actually better for code compren when encountering unfamiliar syntax.

@@

For a parameterized module that takes non-trivial parameters, we need only define the TSM to also take the same parameters and apply them explicitly within the definition of the parameterized module. For example, let us define a parameterized module \lstinline{G : A -> B}:
\begin{lstlisting}[numbers=none,mathescape=|]
signature A = sig { type t; val x : t }
signature B = sig { type u; val y ; u }
syntax $G(M : A)(G : B) at G.u { (* ... *) }
module G(M : A) => mod { 
  type u = M.t; val y = M.x } :> B with syntax $G(M)
\end{lstlisting}
Note that within the definition of \lstinline{G}, type \lstinline{u} does not have a TSL available to it.

@@

To provide the strongest metatheoretic guarantees, we must have that the SL is a total functional language. If we instead defined the SL such that evaluating the parse function may not terminate, this would then cause typechecking to become undecidable.\footnote{A ``hack'' that could be used in practice is to specify a limit to the running time of the parse function. The compiler would enforce this limit to ensure that compilation does not ``hang'' due to an incorrect parser.} Related concerns would arise if the static language supported external (e.g. I/O) effects -- one would generally not want the mere act of parsing and typechecking a program to have an effect on a file system. Mutation effects could be introduced, however, because the static language is kept distinct from the external language. No static values can be ``promoted'' into the external language.

@@

Finally, we have discussed only expression syntax here, but have not considered pattern matching. In fact, we conjecture that it is completely straightforward to extend this work to allow a TSM to define both an expression parser and a pattern parser. Similarly, one could imagine defining TSMs at the level of modules, and at the level of types. We plan to discuss these in the dissertation. Formally specifying these will be left to future work, as it would introduce a rather large amount of technical overhead for minimal conceptual gain. %, the essential ideas are mostly simply conveyed by sticking to formal specifications only at the level of expressions.

@@

\section{Modularly Programmable Type Structure}\label{sec:metamodules}
In the previous section, we assumed that the EL had an otherwise standard, ML-like type structure and considered only TSMs and TSLs as departures from ML. In fact, constructs like case types (shown briefly in use in the previous section), tuples, records and objects are not built primitively into Verse. Instead, they can be expressed using \emph{metamodules} (which are distinct from, though closely related to, {modules}, as we will discuss below). Only  functions are built primitively into the Verse EL.

@@

\subsection{Motivating Example: Labeled Tuples and Regular Strings}\label{sec:metamodules-example}\label{sec:metamodules-motivating-examples}
As a simple introductory example, let us define a \emph{labeled tuple type} for conference papers:
\begin{lstlisting}[numbers=none]
let type Paper = ltuple {SPCT
   title : EPCTrstring /SURL.+EURL/
    SPCTconf : EPCTrstring /SURL([A-Z]+) (\d\d\d\d)EURL/
}
\end{lstlisting}
The \lstinline{let type} construction defines a synonym, \lstinline{Paper}, for a type constructed by applying the \emph{type constructor} (or \emph{tycon}) \lstinline{ltuple} to a \emph{component specification}, which is a  statically valued (cf. Sec. 2) ordered finite mapping from labels to types, written using conventional concrete syntax.

@@

Given a value of type \lstinline{Paper}, we can project out a component value by providing  a  label:% like this:
\begin{lstlisting}[numbers=none]
let val exmpl_conf = # <SNATconfENAT> exmpl 
\end{lstlisting}
Here, \lstinline{#} identifies an \emph{operator constructor} (or \emph{opcon}) parameterized by a statically valued label, written literally here as \lstinline{<SNATconfENAT>}. The resulting \emph{operator}, \lstinline{# <SNATconfENAT>}, is then passed one \emph{argument}, \lstinline{exmpl}. An important point is that this operator is not a function, i.e. it cannot be given function type, because it is able to operate on values of \emph{any} labeled tuple type that has a component labeled \lstinline{conf}, not just \lstinline{Paper}. That type's component specification determines the type of the operation as a whole, here \lstinline{rstring /SURL([A-Z]+) (\d\d\d\d)EURL/}.

@@

We will equip labeled tuple types with an extended set of operators that go beyond the introduction and projection operators just discussed. For example, two labeled tuples can be concatenated (with common components updated with the value on the right) using the \lstinline{ltuple+} operator. Using this operator, we can add a component labeled \lstinline{authors} to \lstinline{exmpl}:
\begin{lstlisting}[numbers=none]
let val a : ltuple {SPCTauthors : EPCTlist(rstring /SURL.+EURL/)} = {["SUSHarry Q. BovikEUS"]}
let val exmpl_paper_final = ltuple+ exmpl a
\end{lstlisting}
We can drop a component using the operator constructor \lstinline{ltuple-}, which is parameterized by a static label:
\begin{lstlisting}[numbers=none]
let val exmpl_paper_anon = ltuple- <SNATauthorsENAT> exmpl_paper_final
\end{lstlisting}
There are many other operators that we might wish to describe, for both labeled tuples and regular strings, but for our purposes it suffices to stop here.

@@

\paragraph{Modules}
It is reasonable to ask if it is possible to express the type structure of any particular labeled tuple type using the module system.  %For simplicity below, let us slightly weaken our insistence on minimalism and build binary products into the EL (and, for now, assume that regular string types are also built in in some suitable manner).%This is not strictly necessary -- we could use a Church-style encoding -- but this simplifies our examples.%then we could try to use the module system to express the type structure of any particular labeled tuple type. For our example above:
For our example type \lstinline{Paper} from above, we might start by defining the following signature:
\begin{lstlisting}
signature PAPER = sig {
  type t
  (* introduction without labels *)
  val intro : title_t -> conf_t -> t
  (* introduction with explicit labels, in both orders *)
  val intro_title_conf : title_t -> conf_t -> t
  val intro_conf_title : conf_t -> title_t -> t
  (* projection *)
  val prj_title : t -> title_t
  val prj_conf : t -> conf_t
}
\end{lstlisting}
Here, we've taken all possible valid introductory and projection operators and called for their expression as functions. Expressing labels explicitly is an important facet of having labeled tuple types in a language, so we must put them into the function identifiers. Even with our minimal EL, we could  implement the semantics of the corresponding operators against this signature using a Church-style encoding (the details are standard and omitted here). Alternatively, if we slightly relax  our insistence on minimalism and add binary products to the EL, we could use them to implement our desired semantics like this:
\begin{lstlisting}
 module Paper : PAPER = mod {
   type t = title_t * conf_t
   fun intro title conf => (title, conf)
   fun intro_title_conf title conf => (title, conf)
   fun intro_conf_title conf title => (title, conf)
   fun prj_title (title, conf) => title 
   fun prj_conf (title, conf) => conf
 }
 \end{lstlisting}
%Compared to the four line type definition at the beginning of Sec.  safe to say that this is an  unacceptable departure from the four line type definition at the beginning of Sec. \ref{sec:metamodules-example}.

@@

Formally, we should be able to derive that $\vdash \texttt{Paper}~\mathtt{type} \leadsto \texttt{string} \times (\texttt{string} \times \mathsf{unit})$. The following rule governs type translation:
\begin{mathpar}
\inferrule[trans-ty]{
    \vdash_\Phi \mathbf{ty}(m\cdot c; \sigma_\text{param}) :: \mathsf{Ty}\\
    \mathbf{ty}(m\cdot c; \sigma_\text{param})~\mathtt{sval}_\Phi\\
	\mathbf{translator}(\Phi; m\cdot c)=\sigma_\text{translator}\\
	\sigma_\text{translator}(\sigma_\text{param}) \Downarrow \sigma_\text{trans}\\
	\sigma_\text{trans} \uparrow_\Phi^m \tau_\text{abstrans} \parallel {\delta} : {\Delta}\\
	{\Delta} \vdash \tau_\text{abstrans}~\mathtt{itype}
}{
	\vdash_\Phi \mathbf{ty}(m.c;\sigma_\text{param})~\mathtt{type}\leadsto[{\delta}]\tau_\text{abstrans}
}
\end{mathpar}
Note that this rule is again slightly simplified here. It would need to be modified to handle external type abstraction, for example. However, it is pedagogically useful to begin with this simpler rule. Note that our purpose in this proposal is not to give a complete formal account of our contributions, but only to sketch out what this formal account will look like.

@@

\subsubsection{Limitations}
The main limitation of the mechanism just described is that it only supports defining types and operators that do not require adding new contexts to the EL typing judgement. No new binding structure can be defined. Fortunately, a number of interesting examples found in existing languages have this character, including labeled tuple types, record types, labeled sum types, object types (of various designs), regular string types and range-restricted numeric types. We plan to detail these in the thesis. On the other hand, constructs like ML-style exceptions, OCaml-style open datatypes or Java-style class-based object systems would not be possible because they require statically tracking which constructors/classes are in scope. We plan to discuss adding support for new user-defined static contexts as an avenue for future work in the dissertation.

@@

To further validate our work, we plan to detail a number of the examples mentioned above, including regular strings, case types, and a simple prototypic object system. We anticipate that the writing involved to do this, along with other writing tasks, will take another \textbf{2 weeks}. So the total for this section is 9 weeks.
% The semantics uses internal type abstraction to guarantee that metamodule providers can reason modularly about translation invariants, just as it uses external type abstraction to guarantee that module providers can reason modularly about representation invariants.%For example, a metamodule that expresses the type structure of records wouldLblTyList introduce a type constructor  \texttt{record} parameterized by a static mapping from labels to types, and two operators: the introductory operator \texttt{intro}, parameterized by a list of row labels, and the row projection operator \texttt{prj}, parameterized by a row label. It might then choose to internally represent records by binary products, which the IL provides. Critically, the choice of internal representation is \emph{computed} based on each record type's parameter, and it is held abstract outside the metamodule using internal type abstraction, facilitating modular reasoning about translation invariants (just as external type abstraction facilitates modular reasoning about representation invariants in the module system). Metamodules like this can express many other constructs, including variants of records that support additional operations, labeled sums, object systems, constrained strings, typechecked operations on format strings like \texttt{printf} and typechecked foreign function interfaces. By using %This simplifies the Verse core language relative to the core languages of ML,  Scala and other comparable languages.

>>>>

\paragraph{Abstract Syntax} The abstract syntax of {patterns}, $p$, over strings, $s$, is specified as below:\[p ::= \textbf{empty} ~|~ \textbf{str}(s) ~|~ \textbf{seq}(p; p) ~|~ \textbf{or}(p; p) ~|~ \textbf{star}(p) ~|~ \textbf{group}(p)\]
One way to express this abstract syntax is by defining a recursive sum type \cite{pfpl}. Verse supports these as datatypes, which are comparable to datatypes in ML (we plan to show how the type structure  of datatypes can in fact be expressed in libraries later):

\begin{lstlisting}[numbers=none]
datatype Pattern {
    Empty | Str of string | Seq of Pattern * Pattern 
    | Or of Pattern * Pattern | Star of Pattern | Group of Pattern
}
\end{lstlisting}

@@

\paragraph{Formalization} To introduce the formal specification of TSMs, let us consider the rule in the EL's semantics for handling TSM application to a delimited form. To simplify matters, we will first consider only unparameterized TSMs (i.e. TSMs defined at a single type, e.g. the  datatype \lstinline{Pattern}, rather than a parameterized family of types): 
\begin{mathpar}
\inferrule[syn-aptsm]{
    \vdash \psi ~@~ \sigma_\text{ty}~\{\sigma_\text{parser}\}\\
    \mathbf{parsestream}(b) = \sigma_\text{ps}\\
    \sigma_\text{parser}(\sigma_\text{ps}) \Downarrow \sigma_\text{exp}\\
    \sigma_\text{exp} \uparrow e_\text{exp}\\\\
    \Gamma; \emptyset \vdash e_\text{exp} \Leftarrow \sigma_\text{ty} \leadsto \iota_\text{trans}
}{
    \Gamma_\text{out}; \Gamma \vdash \mathbf{aptsm}(\psi; b) \Rightarrow \sigma_\text{ty} \leadsto \iota_\text{trans}
}
\end{mathpar}
Looking at the conclusion of the rule, we see that TSM application takes the abstract form $\mathbf{aptsm}(\psi; b)$. The metavariable $\psi$ ranges over \emph{TSM expressions} and $b$ ranges over {bodies} of delimited forms (i.e. the sequences of characters between the delimiters in the examples above). TSM expressions consist only of TSM definitions here (cf. the definition of \lstinline{pattern} above), though when we formalize parameterized TSMs, we will distinguish TSM definitions from TSM expressions, which involve parameter application.%Note that for simplicity in the formal system, we have tacitly assumed that TSM names are drawn from a set that is disjoint from the set of variable names, so that function application and TSM application desugar to the distinct abstract forms. %In practice, we would use a common set of names and include a disambiguation step, but making the assumption of disjoint names allows us to avoid specifying this. %In practice, there would be an additional step that determined whether a function or a TSM was being applied based on the current context,

@@

As another example, we can use TSLs to express derived list syntax. For example, if we use a datatype, we can define the TSL directly upon declaring the list type:
\begin{lstlisting}[numbers=none]
datatype list('a) { Nil | Cons of 'a * list('a) } with syntax {
  static fn (body :: ParseStream) => 
    (* ... comma-delimited spliced exps ... *)
}
\end{lstlisting}
Together with the TSL for patterns, this allows us to write a list of patterns like this:
\begin{lstlisting}[numbers=none]
let val x : list(P.t) = [/SURL\dEURL/SHTML, EHTML/SURL\d\dEURL/SHTML, EHTML/SURL\d\d\dEURL/]
\end{lstlisting}
From the client's perspective, it is essentially as if the language had built in derived syntax for lists and regular expression patterns directly.%However, we did not need to build in this syntax primitively.%The only constraint is that this syntax must be used in an analytic position, which we argue is actually better for code compren when encountering unfamiliar syntax.

@@

For a parameterized module that takes non-trivial parameters, we need only define the TSM to also take the same parameters and apply them explicitly within the definition of the parameterized module. For example, let us define a parameterized module \lstinline{G : A -> B}:
\begin{lstlisting}[numbers=none,mathescape=|]
signature A = sig { type t; val x : t }
signature B = sig { type u; val y : u }
syntax $G(M : A)(G : B) at G.u { (* ... *) }
module G(M : A) => mod { 
  type u = M.t; val y = M.x } :> B with syntax $G(M)
\end{lstlisting}
Note that within the definition of \lstinline{G}, type \lstinline{u} does not have a TSL available to it.

@@

To provide the strongest metatheoretic guarantees, we must have that the SL is a total functional language. If we instead defined the SL such that evaluating the parse function may not terminate, this would then cause typechecking to become undecidable.\footnote{This source of undecidability might not be particularly important in practice, for the same reason that non-termination is tolerated in most major languages.} Related concerns would arise if the static language supported external (e.g. I/O) effects -- one would generally not want the mere act of parsing and typechecking a program to have an effect on a file system. Mutation effects could be introduced, however, because the static language is kept distinct from the external language. No static values can be ``promoted'' into the external language.

@@

We have discussed only expression syntax here, but have not considered pattern matching. In fact, we conjecture that it is completely straightforward to extend this work to allow a TSM to define both an expression parser and a pattern parser. Similarly, one could imagine defining TSMs at the level of modules, and at the level of types. We plan to discuss these in the dissertation. Formally specifying these will be left to future work, as it would introduce a rather large amount of technical overhead for minimal conceptual gain. %, the essential ideas are mostly simply conveyed by sticking to formal specifications only at the level of expressions.

@@

\section{Modularly Programmable Type Structure}\label{sec:metamodules}
In the previous section, we assumed that the EL had an otherwise standard, ML-like type structure and considered only TSMs and TSLs as departures from ML. In fact, constructs like datatypes (shown briefly in use in the previous section), tuples, records and objects are not built primitively into Verse. Instead, they can be expressed using \emph{metamodules} (which are distinct from, though closely related to, {modules}, as we will discuss below). Only  functions are built primitively into the Verse EL.

@@

\subsection{Motivating Example: Labeled Tuples and Regular Strings}\label{sec:metamodules-example}\label{sec:metamodules-motivating-examples}
As a simple introductory example showing the sorts of types and operators we might like to have available to us as clients of a language, let us define a simple \emph{labeled tuple type} for conference papers:
\begin{lstlisting}[numbers=none]
let type Paper = ltuple {SPCT
   title : EPCTrstring /SURL.+EURL/
    SPCTconf : EPCTrstring /SURL([A-Z]+) (\d\d\d\d)EURL/
}
\end{lstlisting}
The \lstinline{let type} construction defines a synonym, \lstinline{Paper}, for a type constructed by applying the \emph{type constructor} (or \emph{tycon}) \lstinline{ltuple} to a \emph{component specification}, which is a  statically valued (cf. Sec. 2) ordered finite mapping from labels to types, written using conventional concrete syntax.

@@

Given a value of type \lstinline{Paper}, we can project out a component value by providing  a  label:% like this:
\begin{lstlisting}[numbers=none]
let val exmpl_conf = # <SNATconfENAT> exmpl 
\end{lstlisting}
Here, \lstinline{#} identifies an \emph{operator constructor} (or \emph{opcon}) parameterized by a statically valued label, written literally here as \lstinline{<SNATconfENAT>}. The resulting \emph{operator}, \lstinline{# <SNATconfENAT>}, is then passed one \emph{argument}, \lstinline{exmpl}. An important point is that this operator is not a function, i.e. it cannot be given function type, because it is able to operate on values of \emph{any} labeled tuple type that has a component labeled \lstinline{conf}, not just \lstinline{Paper}. The argument type's component specification determines the type of the operation as a whole, here \lstinline{rstring /SURL([A-Z]+) (\d\d\d\d)EURL/}.

@@

We will equip labeled tuple types with an extended set of operators that go beyond the introduction and projection operators just discussed. For example, two labeled tuples can be concatenated (with common components updated with the value on the right) using the \lstinline{ltuple+} operator. Using this operator, we can add a component labeled \lstinline{authors} to \lstinline{exmpl}:
\begin{lstlisting}[numbers=none]
let val a : ltuple {SPCTauthors : EPCTlist(rstring /SURL.+EURL/)} = {["SUSHarry Q. BovikEUS"]}
let val exmpl_paper_final = ltuple+ exmpl a
\end{lstlisting}
We can drop a component using the operator constructor \lstinline{ltuple-}, which is parameterized by a static label:
\begin{lstlisting}[numbers=none]
let val exmpl_paper_anon = ltuple- <SNATauthorsENAT> exmpl_paper_final
\end{lstlisting}
There are many other operators that we might wish to describe, for both labeled tuples \cite{Cardelli91operationson} and regular strings \cite{sanitation-psp14}, but for our purposes it suffices to stop here.

@@

\paragraph{Modules}
It is reasonable to ask if it is possible to express the type structure of any particular labeled tuple type using the module system.  %For simplicity below, let us slightly weaken our insistence on minimalism and build binary products into the EL (and, for now, assume that regular string types are also built in in some suitable manner).%This is not strictly necessary -- we could use a Church-style encoding -- but this simplifies our examples.%then we could try to use the module system to express the type structure of any particular labeled tuple type. For our example above:
For our example type \lstinline{Paper} from above, we might start by defining the following signature:
\begin{lstlisting}
signature PAPER = sig {
  type t
  (* introduction without labels *)
  val intro : title_t -> conf_t -> t
  (* introduction with explicit labels, in primary order *)
  val intro_title_conf : title_t -> conf_t -> t
  (* projection *)
  val prj_title : t -> title_t
  val prj_conf : t -> conf_t
}
\end{lstlisting}
Here, we've taken all possible valid introductory and projection operators and called for their expression as functions. Expressing labels explicitly is an important facet of having labeled tuple types in a language, so we put them into the function identifiers. This is, however, suboptimal because in client code, the argument values will not appear adjacent to the corresponding labels.

Even with our minimal EL, we could  implement the semantics of the corresponding operators against this signature using a Church-style encoding (the details are standard and omitted here). Alternatively, if we slightly relax  our insistence on minimalism and add binary products to the EL, we could use them to implement our desired semantics like this:
\begin{lstlisting}
 module Paper : PAPER = mod {
   type t = title_t * conf_t
   fun intro title conf => (title, conf)
   fun intro_title_conf title conf => (title, conf)
   fun prj_title (title, conf) => title 
   fun prj_conf (title, conf) => conf
 }
 \end{lstlisting}
%Compared to the four line type definition at the beginning of Sec.  safe to say that this is an  unacceptable departure from the four line type definition at the beginning of Sec. \ref{sec:metamodules-example}.

@@

Formally, we should be able to derive that $\vdash \texttt{Paper}~\mathtt{type} \leadsto \texttt{string} \times (\texttt{string} \times \mathsf{unit})$. The following rule governs type translation:
\begin{mathpar}
\inferrule[trans-ty]{
    \vdash_\Phi \mathbf{ty}(m\cdot c; \sigma_\text{param}) :: \mathsf{Ty}\\
    \mathbf{ty}(m\cdot c; \sigma_\text{param})~\mathtt{sval}_\Phi\\
	\mathbf{translator}(\Phi; m\cdot c)=\sigma_\text{translator}\\
	\sigma_\text{translator}(\sigma_\text{param}) \Downarrow \sigma_\text{trans}\\
	\sigma_\text{trans} \uparrow_\Phi^m \tau_\text{abstrans} \parallel {\delta} : {\Delta}\\
	{\Delta} \vdash \tau_\text{abstrans}~\mathtt{itype}
}{
	\vdash_\Phi \mathbf{ty}(m\cdot c;\sigma_\text{param})~\mathtt{type}\leadsto[{\delta}]\tau_\text{abstrans}
}
\end{mathpar}
Note that this rule is again slightly simplified here. It would need to be modified to handle external type abstraction, for example. However, it is pedagogically useful to begin with this simpler rule. Note that our purpose in this proposal is not to give a complete formal account of our contributions, but only to sketch out what this formal account will look like.

@@

\subsubsection{Limitations}
The main limitation of the mechanism just described is that it only supports defining types and operators that do not require adding new contexts to the EL typing judgement. No new binding structure can be defined. Fortunately, a number of interesting examples found in existing languages have this character, including labeled tuple types, record types, labeled sum types, object types (of various designs), regular string types and range-restricted numeric types. We plan to detail these in the thesis. On the other hand, constructs like ML-style exception values, OCaml-style open datatypes or Java-style class-based object systems would not be possible because they require statically tracking which constructors/classes are in scope. We plan to discuss adding support for new user-defined static contexts as an avenue for future work in the dissertation.

@@

To further validate our work, we plan to detail a number of the examples mentioned above, including regular strings, datatypes, and a simple prototypic object system. We anticipate that the writing involved to do this, along with other writing tasks, will take another \textbf{2 weeks}. So the total for this section is 9 weeks.
% The semantics uses internal type abstraction to guarantee that metamodule providers can reason modularly about translation invariants, just as it uses external type abstraction to guarantee that module providers can reason modularly about representation invariants.%For example, a metamodule that expresses the type structure of records wouldLblTyList introduce a type constructor  \texttt{record} parameterized by a static mapping from labels to types, and two operators: the introductory operator \texttt{intro}, parameterized by a list of row labels, and the row projection operator \texttt{prj}, parameterized by a row label. It might then choose to internally represent records by binary products, which the IL provides. Critically, the choice of internal representation is \emph{computed} based on each record type's parameter, and it is held abstract outside the metamodule using internal type abstraction, facilitating modular reasoning about translation invariants (just as external type abstraction facilitates modular reasoning about representation invariants in the module system). Metamodules like this can express many other constructs, including variants of records that support additional operations, labeled sums, object systems, constrained strings, typechecked operations on format strings like \texttt{printf} and typechecked foreign function interfaces. By using %This simplifies the Verse core language relative to the core languages of ML,  Scala and other comparable languages.
